---
title: "PSY260 - Research Methods and Analysis"
subtitle: "R Workshop 1 - Using R for Data Processing Questions"
output: 
  html_document:
    css: "../../style.css"
    toc: true
    toc_float: true
    theme: united
    highlight: tango
author: "Dr Glenn Williams"
date: "2020-09-19 (Updated `r Sys.Date()`)"
---

Make sure this file is in the same folder as your RProject. Open it up and attempt the questions below.

# Question 1

What's wrong with the following code? Fix it in the second block below.

```{r question_1, eval = FALSE}
sequence < - c(1, 2, 3, 4, 5)
```

Can you explain why the code above does not work?

**A: It has a space between < and - so assignment won't work (it will read this as less than then minus, but fail because sequence doesn't exist)**

```{r answer-1}
sequence <- c(1, 2, 3, 4, 5)
```

# Question 2

Why doesn't this return anything? Change the code to make it run.

```{r question-2-a}
# draw 10 times from a uniform distribution between 0 and 1
my_object <- runif(n = 10, min = 0, max = 1)
```

```{r question-2-b, eval = FALSE}
my_obejct
```

Can you explain why the code above does not work?

**A: The name of the object has a typo**

```{r answer-2}
my_object
```

# Question 3

Create an object with names of 5 of your closest friends, family members, or favourite musicians. Name this object "people". *Hint* you will need to concatenate 5 strings with the `c()` function.

```{r answer-3}
people <- c("Rachel", "Ross", "Joey", "Chandler", "Monica")
```

# Question 4

Create an object called `scores` that will store the scores of your 5 people on some test. Give them some random numbers. You can do this by hand, or learn how to use the `rnorm()` function by typing `?rnorm()` into the console.

```{r answer-4}
scores <- rnorm(n = 5, mean = 100, sd = 15)
```

# Question 5

Finally, create some ages for your people and store them in the object `ages`. Feel free to make up their ages. If you want to learn a new function, look into how `sample()` works. This is a function that draws numbers randomly from a set of numbers. Again, type `?sample()` to figure this out.

```{r answer-5}
ages <- sample(20:30, size = 5, replace = TRUE)
```

# Question 6

Create a data frame called `my_data` consisting of your people, their ages, and test scores. Remember to use the `data.frame()` function here. Again, if you'd like some help on how this works, type `?data.frame()` into the console.

```{r answer-6}
my_data <- data.frame(
  people = people,
  ages = ages,
  scores = scores
)

my_data
```

# Question 7

Change the name of the first person in `my_data` to someone else. Don't worry about changing their other values. *Hint*: You will need to index that particular value in your data.frame and assign it a new value.

```{r answer-7}
my_data[1, 1] <- "Phoebe"
my_data
```

# Question 8

Access the 3rd row and 2nd column of `my_data`.

```{r answer-8}
my_data[3, 2]
```

# Question 9

Access the 2nd column all rows of `my_data`.

```{r answer-9}
my_data[, 2]
```

# Question 10

Calculate the mean age of the people in your sample. To do this, use the function `mean()`.

```{r answer_10}
mean(my_data$ages)
```

# Question 11

Calculate the standard deviation for the age of the people in your sample. To do this, use the function `sd()`.

```{r answer_11}
sd(my_data$ages)
```

# Question 12

You can chain together functions in R. 

One function we can use is `unique()`, which returns a vector or data frame based on what we give it, only with duplicates removed. Another function is 'length()` which tells us how many things there are in a vector.

Combine these two functions to get the number of unique names in your `people` column in `my_data`. *Hint*: pass your column to your functions. One function will be around the other one, e.g. `x(y(...))`.

```{r answer_12}
length(unique(my_data$people))
```

That's it for the introductory section. Well done on completing your first class in R!
