---
title: "PSY260 - Research Methods and Analysis"
subtitle: "R Workshop 1 - Using R for Data Processing Questions"
output: 
  html_document:
    css: "../../style.css"
    toc: true
    toc_float: true
    theme: united
    highlight: tango
author: "Dr Glenn Williams"
date: "2021-10-25 (Updated `r Sys.Date()`)"
---

Now you know how to download and install R and RStudio, how to set up RStudio Projects, and how to create and use a notebook, you're ready to get started working within a notebook in RStudio.

Complete the following exercises. This will get you used to working with R notebooks and carrying out basic operations. After this, complete the questions at the end of the session or in your own time.

# Exercises

Please complete the following exercises within this file (questions.Rmd). 

## Exercise 1

Make a code chunk. In this, add together the numbers 1, 5, 999, and 67819.

## Exercise 2

Make a code chunk. In this, divide 1825031 by 534.

## Exercise 3

Make a code chunk. Perform a boolean operation to check whether "Dog" is equal to "Cat".

## Exercise 4

Make a code chunk. Concatenate the string "I'm using R" with the number 7. What output do you get? Can you explain why it is presented in this way?

## Exercise 5

Make a code chunk. In this, calculate the mean of the numbers 1 through to 10 (using `mean()`). Also in this chunk, calculate the median of the numbers 1 through to 10 (using `median()`). Also in this chunk, calculate the standard deviation of the numbers 1 through to 10 (using `sd()`). 

Can you simplify your code to cut down on repetition? 

*Note*: If you don't get the numbers you'd expect from e.g. `mean()`, make sure you're concatenating your values with `c()`. If you don't, it will just return the mean of the first value.

## Exercise 6

Make a code chunk. Create an object called "numbers" that has the numbers 1, 5, 10, and 267 in it. Access the final number. After this, replace the final number with 3.

## Exercise 7

Replace the text in your R markdown file to explain what you did with your data. It's usually best to explain what you did at each step before each code chunk.

# Questions

These questions are designed to test your knowledge further in R. Please try to tackle these once you're done with the exercises.

## Question 1

What's wrong with the following code? Fix it in the second block below.

```{r question_1, eval = FALSE}
sequence < - c(1, 2, 3, 4, 5)
```

Can you explain why the code above does not work?

```{r answer-1}

```

## Question 2

This code generates random numbers between 0 and 1 from a uniform distribution.

```{r question-2-a}
# draw 10 times from a uniform distribution between 0 and 1
my_object <- runif(n = 10, min = 0, max = 1)
```

Why doesn't the below code run?

```{r question-2-b, eval = FALSE}
my_obejct
```

Can you explain why the code above does not work? Write your answer below and the valid R code needed to fix this issue.

```{r answer-2}

```

## Question 3

Create an object with names of 5 of your closest friends, family members, or favourite musicians. Name this object "people". *Hint* you will need to concatenate 5 strings with the `c()` function.

```{r answer-3}

```

## Question 4

Create an object called `scores` that will store the scores of your 5 people on some test. Give them some random numbers. You can do this by hand, or learn how to use the `rnorm()` function by typing `?rnorm()` into the console.

```{r answer-4}

```

## Question 5

Finally, create some ages for your people and store them in the object `ages`. Feel free to make up their ages. If you want to learn a new function, look into how `sample()` works. This is a function that draws numbers randomly from a set of numbers. Again, type `?sample()` in the console and press enter to figure this out.

```{r answer-5}

```

## Question 6

Create a data frame called `my_data` consisting of your people, their ages, and test scores. Remember to use the `data.frame()` function here. Again, if you'd like some help on how this works, type `?data.frame()` into the console and press enter.

```{r answer-6}

```

## Question 7

Change the name of the first person in `my_data` to someone else. Don't worry about changing their other values. *Hint*: You will need to index that particular value in your data.frame and assign it a new value.

```{r answer-7}

```

## Question 8

Access the 3rd row and 2nd column of `my_data`.

```{r answer-8}

```

## Question 9

Access the 2nd column all rows of `my_data`.

```{r answer-9}

```

## Question 10

Calculate the mean age of the people in your sample. To do this, use the function `mean()`.

```{r answer-10}

```

## Question 11

Calculate the standard deviation for the age of the people in your sample. To do this, use the function `sd()`.

```{r answer-11}

```

## Question 12

You can chain together functions in R. 

One function we can use is `unique()`, which returns a vector or data frame based on what we give it, only with duplicates removed. Another function is `length()` which tells us how many things there are in a vector.

Combine these two functions to get the number of unique names in your `people` column in `my_data`. 

*Hint*: pass your column to your functions. One function will be around the other one, e.g. `x(y(...))`.

```{r answer-12}

```

That's it for the introductory section. Well done on completing your first class in R!
