---
title: "Using R for Data Processing"
subtitle: "The What and Why of R"
author: "Glenn Williams"
institute: "University of Sunderland"
date: "2021-10-25 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
set.seed(100)
library(here)
```

```{r functions, include = FALSE}
colorise <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}
```

# What is R?

.pull-left[

- **Statistical programming language** used for processing, summarising, analysing, and graphing data.

- **Free and open source**, so anyone can see how it works and add to the development of the program.

- Packages and changes are **vetted by reviewers**, so we can be sure most things we do in R are appropriate.

- Often used with **RStudio** which makes working with **notebooks**, **projects**, and **version control** easier. More on these features later.

]

.pull-right[

```{r r-logos, echo = FALSE, out.width = 300, fig.align = "center"}
knitr::include_graphics(c(
   "../../img/R-logo.png",
    "../../img/rstudio-logo.png"
))
```

]

---
# Why Should I Care?

.pull-left[
- Free software means **anyone can use it at no cost**. This makes science more inclusive, and allows for easier confirmation of analyses.

- Programming languages force you to **document all decisions** made with the data. This makes your research more **transparent**.
]

.pull-right[
```{r anchorman, echo = FALSE, out.width = 350, fig.align = "center", fig.cap="You supporting Open Science"}
knitr::include_graphics(c(
    "../../img/anchorman.jpg"
))
```
]

- With **Open Source** Software we can see how the program works and improve/extend it where needed. **Finding and fixing problems is easier and quicker**.

- Free, Open Source software is crucial for **Open Science**.

---
# Sharing is Caring

- Is your reseach **replicable**? If I follow your methods with **new participants**, will I get **similar results**?

- Is your research **reproducible**? If I follow your methods with **your data**, will I get **the same results**?

- Checking these things is made possible if you share your materials, data, and steps for analysis.

- Sites like the [Open Science Foundation (OSF)](https://osf.io) and [GitHub](https://github.com/) allow us to host our research products online. They also track changes to your work.

- Opening up your research can be scary, but **it'll probably make you more careful**, allow people to build on your work, and allow science to be more reliable and able to self-correct.

---
# Why Should I Care?

- Veldkamp et al. (2014): 63% of articles contained **at least one *p*-value that was incorrect**. In 20.5% of cases, these errors led to **erroneous decisions** about the statistical significance of an effect.

- How does this happen? 

    - Not documenting your methods fully makes **detecting mistakes much more difficult**.
    - Transcribing results from visual interfaces like SPSS means you introduce **human error**.
    
- How can we fix this problem? **Notebooks**!

    - You never write the results, just the methods for making them.
    - Every step of your analysis is documented.
    - **If your data changes, your results are instantly updated**.

---
# Still Not Convinced?

.pull-left[
- Easy to learn is not Easy to use. 
    
    - How do you **filter observations** in Excel? In R, use `filter()`
    - How do you do a **t-test** in SPSS? In R, use `t.test()`
    - Learning to code might take longer, but implementing it is often easier down the line. 
    
- **You won't document everything you do in Excel**. That's a problem if you notice a mistake.
- **R scales up**. Imagine processing 1,000,000 rows of data in Excel. R can do that as easily as 10 rows.
]
    
.pull-right[
```{r excel, echo = FALSE, out.width = 300, fig.align = "center"}
knitr::include_graphics(c(
    "../../img/excel-problems.png",
    "../../img/pikachu.png"
))
```
]

---
# Example Notebook

```{r r-codebook-example, echo = FALSE, out.width = 500, fig.align = "center"}
knitr::include_graphics(c(
   "../../img/rnotebook-example.png"
))
```

Software versions are easy to report, and we can **see both the code to produce results as well as the results**.

---
# How Do I Share My Work?

- Using R without sharing will help to solve many issues with reproducibility, but making it accessible to others is even better.

.pull-left[
- [The OSF](https://osf.io) is the most user-friendly method of sharing all your research.

    - Servers paid for 50 years, with servers in Europe.
    - **Drag and drop** to upload files.
    - Automatically **tracks updates** to files.
    - Can link **pre-registrations, licenses, and pre-prints** to your project.
]

.pull-right[
```{r the-osf, echo = FALSE, out.width = 300, fig.align = "center", fig.cap="The OSF"}
knitr::include_graphics(c(
   "../../img/osf-screenshot.png"
))
```
]

- [GitHub](https://github.com/) is another option, which has more advanced **version control**, but can be trickier to use.

---
# What We'll Do

After these lessons, you will have:

- used [rstudio.cloud](https://rstudio.cloud/) to learn the basics of how R works.
- created your first notebook.
- processed some raw data and cleaned it up.
- made summaries of our data.
- made graphs of data.
- shared all of this on [The OSF](https://osf.io).

It can be difficult at first, but it **will make things easier** when it comes to the assessment. Don't worry:

- You will struggle to remember the code, but **that's fine**.

- It's standard in most sciences, so if you have a question/issue, it's already been asked and answered somewhere. **Google is your friend**.

- You'll learn a valuable, **transferable skill**.
