---
title: "PSY260 - Research Methods and Analysis"
subtitle: "R Workshop 2 - Managing Data in R"
output: 
  html_document:
    css: "../../style.css"
    toc: true
    toc_float: true
    theme: united
    highlight: tango
author: "Dr Glenn Williams"
date: "2020-10-13 (Updated `r Sys.Date()`)"
---

# Getting Started in R

Please make sure that you have watched the three [introductory videos on Using R for Data Processing](https://canvas.sunderland.ac.uk/courses/40018/pages/lecture-4-using-r-for-data-processing). These cover why you might use R, how to install and create notebooks in R for data processing, and how to do basic data manipulation in R.

Seminar 1 covers this content in detail, so please also refer to that lecture if you haven't gotten started in R yet.

# Getting Started

## Assuming you haven't got an RProject in a folder called "my-r-work"

**PLEASE SEE SEMINAR ONE FOR DETAILS.**

Open RStudio. In RStudio, click **File $\rightarrow$ New Project...**. You should see a window like this with some other options. Click on the **New Directory** button:

```{r project-directory, echo = FALSE, out.width = 750, fig.align = "center", fig.cap="creating a new project"}
knitr::include_graphics(c(
   "../../img/project-directory.png"
))
```

Click **New Project** and then set a **Directory Name** for your folder (e.g. "my-r-work"). 

Be sure to save this folder somewhere where you can have access to it after class, e.g. on your Desktop or on a USB.

This will create a folder wherever you have it set to in **Create project as a subdirectory of:**. 

Below, this is just on my Desktop, so this will make a folder called "my-r-work" on my Desktop. Inside this folder, there will be a file with the folder name ending in **.Rproj**.

**IMPORTANT: We will work out of this folder each week in seminars, so save all your work here and make sure you don't lose it.**

```{r project-project, echo = FALSE, out.width = 750, fig.align = "center", fig.cap="creating a new project"}
knitr::include_graphics(c(
   "../../img/r-project.png"
))
```

## Assuming you already have an RProject set up in the folder named "my-r-work"

Double click your project called **my-r-work.RProj** in the folder called "my-r-work".

You have now opened RStudio up from your project. Confirm whether this is the case by looking in the very top right of RStudio. Does it have a Project logo with the name "my-r-work"?

if so, we can be sure all the file paths will work now. If not, be sure you're working from your RProject by following the steps above.

## Creating a Notebook

Click File $\rightarrow$ New File $\rightarrow$ R Notebook. This will open an R notebook in the Editor. It should look like this below (but maybe not in the darkmode I use):

```{r rstudio-interface, echo = FALSE, out.width = 500, fig.align = "center", fig.cap="RStudio Interface"}
knitr::include_graphics(c(
   "../../img/rstudio-interface.png"
))
```

Save this notebook as "seminar-two.Rmd" in the "my-r-work" folder.

# Exercises

Copy and paste the data called **seminar_two_data.csv** from the "psy260-r" folder to your folder called "my-r-work".

This data is made up, but we can assume it looks at how quickly people can respond to a target on screen by pressing a button after having had caffeine or not, and when responding with their **dominant or non-dominant hands**. The dependent variable is thus reaction time in milliseconds and the indipendent variables are caffeine and response hand. We also have ages for our participants.

Now you've opened up RStudio project and notebook, we can get started.

Load the data called "seminar-two-data.csv" into this R session as follows. To do this, create an R code chunk and copy and paste the following code into it.

```{r getting-started, eval = FALSE}
library(tidyverse)
library(here)

raw_data <- read_csv(here("seminar_two_data.csv"))
```

Here, we'll get used to performing some basic data manipulation exercises in R.

## Exercise 1

Make a code chunk. Print out the beginning of the data set either by just typing its name or using `glimpse(raw_data)`. Which columns can you identify as being initially problematic? Are there any?

## Exercise 2

Make a code chunk. Check the unique values for each variable in your data set. Are there any issues with the data set? If so, make a note of them.

## Exercise 5

Make a code chunk. Remove the age data from the data set. Save this new data set as "anonymised_data" within R. 

**Remember to assign this new data to an object in R. Name this anon_data**

## Exercise 4

Make a code chunk. Using your data from Exercise 3, clean up any issues you might have with the caffeine variable. Remember, you can use `mutate()` and `case_when()` together to solve this.

**Assign this to a new object called cleaned_data_one**

If you're unsure how these functions work, type `?mutate` and then `?case_when` in the console. The Help pane outlines how these functions work, and gives examples of working code.

If you're still stuck, refer back to Lecture 2 part 3.

## Exercise 5

Make a code chunk. Using your data from Exercise 4, clean up any issues you might have with the response variable using the same methods outlined in Exercise 3.

**Assign this to a new object called cleaned_data_two**

## Exercise 6

Make a code chunk. Add the age data from your initial data (raw_data) back in to the cleaned data (cleaned_data_two).

## Exercise 7

Make a code chunk. Filter out any observations on the dependent variable which are NAs. Remember, to use `is.na()`. Assign this new data to a object called "cleaned_data_three".

# Questions

Download the questions file from Canvas and place this file in your folder. This file is an R markdown file already populated with some R code chunks. Follow the instructions to test your knowledge of R.
