---
title: "Summarising and Plotting Data in R"
subtitle: "Summarising Data"
author: "Glenn Williams"
institute: "University of Sunderland"
date: "2020-09-29 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
set.seed(100)
library(tidyverse)
library(here)

options(tibble.print_min = 4)
```

```{r functions, include = FALSE}
colorise <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}
```

# Summarising Data

Individual data points are some abstraction away from reality, essentially removing some detail.

- Even with trial level data, we will summarise someone's responses in terms of reaction times or correct responses.

- This removes a lot of the details of the real experience, but allows us to see patterns.

- Similarly, we can summarise trial level data in a number of ways to help communicate a pattern or idea.

We can present summaries of data as a plot, or even as a table of descriptive statistics.

---
# Descriptive Statistics

We can often summarise data from individual trials in terms of measures of **central tendency** and **dispersion**.

- **Central tendency**: What is the average score? This can be determined in a few ways.

    - Mean: used to describe data that are normally distributed. Can be misleading when outliers are present.

    - Median: used to describe data that are skewed in some way away from a normal distribution. Suppresses the impact of outliers.

- **Dispersion**: What is the spread of scores (e.g. for individuals) around the average score?

    - Standard Deviation: the spread of the data from the sample mean. With a normal distribution approximately 1, 2, and 3 SDs capture approximately 68, 95, and 99.7% of the data.

    - Interquartile Range: Rank orders data into four equal parts. We get 3 values: the middle part of the first region – or quartile – (i.e. 25%), the median (i.e. 50%), and the middle part of the third quartile (i.e. 75%). Used for non-normal data.

---
# Why Make these in R?

---
# Summarise


---
# Summarise


---
# Grouping


---
# Group by and Summarise
