---
title: "PSY260 - Research Methods and Analysis"
subtitle: "R Workshop 3 - Summarising and Plotting Data in R"
output: 
  html_document:
    css: "../../style.css"
    toc: true
    toc_float: true
    theme: united
    highlight: tango
author: "Dr Glenn Williams"
date: "2020-10-19 (Updated `r Sys.Date()`)"
---

Make sure this file, **03_questions.Rmd**, is in the same folder as your RProject. 
Create a folder here called "data" (if you haven't already done so) and copy the data file "factorial_data.csv" from the data folder in the "psy260-r" folder to this location.

Open up **03_questions.Rmd** in your folder up and attempt the questions below. 

This set of questions will prompt you to summarise a messy data set from beginning to end.

```{r setup, message = FALSE}
# load packages
library(tidyverse)
library(here)

# load data
qdata <- read_csv(here("data", "cleanish_data.csv"))

# inspect data
qdata
```

This data set is that from lecture 2b - Data Checking in R. The columns have the following interpretation:

- **X1** = Row number for the data.

- **subj_id** = Subject/participant ID for the person who gave the data to us.

- **list_id** = list ID for how items were randomised.

- **item_id** = Item ID.

- **variable A** = Implied sentence location (upwards or downwards); e.g. "the bird flew in the sky." a1 is up, a2 is down.

- **variable B** = Location of target (up or down on the screen). b1 is up, b2 is down.

# Question 1

Remove any unnecessary columns from this data set.

```{r question-1}

```

# Question 2

Give proper levels names to the levels A1 and A2 in column A and to B1 and B2 in column B.

```{r question-2}

```

# Question 3

Give proper (column) names to the variables A, B, and Y.

```{r answer-3}

```

# Question 4

Make a summary of the data creating averages of the reaction times for each level of each condition for every participant. Assign this to the object `summary_data`.

```{r answer-4}

```

*Hint:* No advanced logic is needed here. Just use `rep(min:max, each = n)`. To see how this function works, use `?rep`.

# Question 5

Remove participants 1 through to 10 from `summary_data`.

```{r answer-5}

```

# Question 6

Make a boxplot of the data. Does there appear to be any reliable differences between conditions? Explain your logic.

```{r answer-6}

```

# Question 7 (Advanced, Bonus Material)

We can perform inferential statistics in R. 

I have simulated some data below for a 2 group experiment. The two groups are "control" and "treatment" which are assigned to the column "condition" while the outcome is in the column "outcome". (It doesn't matter what this column or study represents, data only have meaning we assign to it.)

```{r simulate-data-question-7}
n <- 40

q8dat <- tibble(
  condition = rep(c("control", "treatment"), times = n),
  outcome = c(
    rnorm(n, mean = 400, sd = 100),
    rnorm(n, mean = 500, sd = 150)
  )
)
```

We can analyse this data using a t-test as follows:

```{r t-test-question-7}
t.test(outcome ~ condition, data = q8dat)
```
We get the above summary of the t-test. (Your exact values may vary due to simulating this data.)

We can see above that we can analyse data using a formula interface `outcome ~ treatment`. This means scores to the left of the `~` are predicted by the values assigned to the right side of the `~`; i.e. your condition dictates your score.

We can similarly perform an ANOVA using this method. This means we perform an ANOVA with two main effects (e.g. a two-way ANOVA) as follows:

```{r anova-example-1, eval = FALSE}
aov(outcome ~ FACTOR_A + FACTOR_B, data = YOUR_DATA)
```

Here, any terms added to the right of the `~` are evaluated as main effects if we make their effects additive. If instead we use a multiplication symbol, we get both main effects **and interactions**.

Perform an ANOVA for your data. To get your typical ANOVA table output, you must use `summary(aov(...))` or assign your ANOVA model to an object and call `summary()` on that object.

```{r answer-7-1}

```

This answer may appear correct, but with repeated measures we need to tell R to adjust the errors (i.e. to conduct a within- rather than a between-subjects ANOVA) for any terms we want to estimate within each participant. Do this as follows:

```{r anova-example-2, eval = FALSE}
aov(
  outcome ~ FACTOR_A * FACTOR_B + 
    Error(SUBJECT_ID_COLUMN/(FACTOR_A * FACTOR_B)), 
  data = YOUR_DATA
)
```

Adjust your code to incorporate this. Do your effects change? Can you explain why?

```{r answer-7-2}

```
